---
title: "Rescaled version"
author: "Mehdi BOUDAOUD"
date: "2024-07-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import libraries

```{r}
library(readxl)
library(dplyr)
library(data.table)
```

## Load data

```{r}
## Load Stations names 
X92st_list <- read_excel("Data/92st_list.xlsx")
X92st_list$St_Name[c(65,66,67)] <- c(
  "QUEBEC_JEAN LESAGE INTL A",
  "MONTREAL_PIERRE ELLIOTT TRUDEAU INTL A",
  "MONTREAL_MIRABEL INT'L A"
)
StationsNames <- X92st_list$St_Name
Data_list <- list()
for (St in StationsNames) {
  print(paste0(
    "Loading : ", St
  ))
  
  ## Assign the path
  St_path <- "Data/92st_RData/" |> 
    paste0(St, ".RData")
  
  ## Load df
  load(St_path)
  
  ## Filter data 
  df$Year <- df$Date |> 
    format("%Y")
  df <- df |> 
    dplyr::filter(Temp > -100) |> 
    dplyr::filter(Year > 1980) |> 
    dplyr::filter(Year < 2012)
  
  ## Save the data in a list
  Data_list[[St]] <- df
}
```


## Define the rescale function

This function takes a data frame of station as input and return a list containing the re-scaled data, the minimum and maximum values.
```{r}
rescale <- function(df){
  df_min <- df$Temp |> min()
  df_max <- df$Temp |> max()
  
  df$Temp <- 
    (df$Temp - min(df$Temp))/(max(df$Temp) - min(df$Temp))
  
  list(
    "Data" = df,
    "Max" = df_max,
    "Min" = df_min
  )
}
```


## Calculate the bandwith minimum to have a certain number of modes 


```{r}
MAX_MODES <- 5

bandwiths <- matrix(
  ncol = MAX_MODES + 1,
  nrow = 92
) |> as.data.frame()

names(bandwiths) <- 
  c(
    "Station",
    paste0(seq(1, MAX_MODES), " pic(s)")
  )
bandwiths$Station <- names(Data_list)
for (st in bandwiths$Station) {
  print('=========================')
  print(st)
  print('=========================')
  
  df_rescaled_list <- 
      Data_list[[st]] |> 
      rescale()
  
  for (i in 1:MAX_MODES) {
    bw <- 
      df_rescaled_list$Data$Temp |> 
      multimode::bw.crit(mod0 = i)
    
    bandwiths[bandwiths$Station == st,][,i+1] <- bw
  }
}
```


## Calculate the gabs of bandwiths
It is clear that we can't find a bandwith that applies to all the stations, for 
the purpose of comparaison we'll find an interval that can apply to all or most of the stations. 

We begin by calculating the gabs between bandwiths for every station.

```{r}

get_bw_gap <- function(df, bw_start = .5){
  df_temp <- df
  for (i in 1:(ncol(df) - 1)) {
    df_temp[paste0(i, " pic(s)")] <-
      bw_start - df[[paste0(i, " pic(s)")]]
  }
  bws_gap_df <- df

  bws_gap_df$`1 pic(s)` <-
    bw_start - df$`1 pic(s)`
  
  for (st in bws_gap_df$Station) {
    for (i in 2:(ncol(df) - 1)) {
      x <- df_temp[df_temp$Station == st, i+1]
      if(x < 0) {
        bws_gap_df[bws_gap_df$Station == st, i+1] <- 0
      } else {
        bws_gap_df[bws_gap_df$Station == st, i+1] <-
          df[df$Station == st, i] -
          df[df$Station == st, i+1]
      }
    }
  }
  # for (i in 2:(ncol(df) - 1)) {
  #   bws_gap_df[paste0(i, " pic(s)")] <-
  #     as.numeric(as.vector(
  #       df[[paste0(i-1, " pic(s)")]]
  #     )) - as.numeric(as.vector(
  #       df[[paste0(i, " pic(s)")]]
  #     ))
  # }
  bws_gap_df[["0 pic(s)"]] <- ifelse(
    bws_gap_df[,ncol(df)] < 0,
    1,
    0
  )
  bws_gap_df[,c(1, MAX_MODES+2, seq(2, MAX_MODES+1))]
  # df_temp
}

opt_n_modes <- function(gap){
  df <- gap |> 
    dplyr::select(Station) |> 
    dplyr::mutate(N_modes = 0)
  
  for (st in gap$Station) {
    df[df$Station == st,
       'N_modes'] <- 
      gap[gap$Station == st, -1] |> 
      # abs() |>
      as.numeric() |> 
      which.max()
      
  }
  df$N_modes <- (df$N_modes - 1) |> 
    factor(levels = seq(0, MAX_MODES))
  return(df)
}

bandwiths |> 
  get_bw_gap(bw_start = 0) |> 
  opt_n_modes()
# x[x$Station == st, -1]
# x[x$Station == st, -1] |> as.numeric() |> which.max()
```

```{r}
N_modes_df <- 
  matrix(
  ncol = MAX_MODES + 2,
  nrow = length(seq(0.02, 0.4, 0.0001))
) |> as.data.frame()

names(N_modes_df) <- 
  c(
    "Starting bw",
    paste0(seq(0, MAX_MODES), " pic(s)")
  )


N_modes_df$`Starting bw` <- seq(0.02, 0.4, 0.0001)
for (i in N_modes_df$`Starting bw`) {
  print(i)
  x <- bandwiths |> 
    get_bw_gap(bw_start = i) |> 
    opt_n_modes() |> 
    dplyr::select(N_modes)
  levels(x$N_modes) <- seq(0, MAX_MODES)
  
  
  N_modes_df[
      N_modes_df$`Starting bw` == i, -1
      # c('One mode', 'Two modes', 'Three modes')
    ] <- x |> table()
}
N_modes_df_long <- N_modes_df |> 
  melt(id.vars = c("Starting bw"),
  variable.name = "N_pics")
colnames(N_modes_df_long) <- c("Starting bw", "Number of pics", "Number of stations")
```
```{r}
N_modes_df_long[N_modes_df_long$`Number of pics` == "0 pic(s)",]

idx <- which(N_modes_df_long$`Number of pics` == "0 pic(s)")
N_modes_df_long <- N_modes_df_long[-idx,]
```

```{r}
library(ggplot2)
fig <- 
  N_modes_df_long |> 
  ggplot(aes(`Starting bw`, `Number of stations`)) +
  geom_line(aes(color = `Number of pics`)) +
  theme_classic() +
  scale_color_brewer(palette = "Dark2") +
  labs(
    x = "Starting bundwith",
    y = "Number of stations",
    color = "Number of pics"
  )
fig
ggsave("Figures/fig308.pdf", width = 7, height = 5)
```


```{r}
library(plotly)
ggplotly(fig)
```

## Save
This script save the last graphic into the sepecified path
```{r}
library(ggpubr)

path <- "Figures/fig.pdf"
ggsave(filename = path, width = 7, height = 5)
```

```{r}
N_modes_df_long[
  N_modes_df_long$`Starting bw` == 0.028,
]

x <- bandwiths |> 
    get_bw_gap(bw_start = 0.0288) |>
    opt_n_modes()
x[x$N_modes == 2,]
x$N_modes |> table()
```

```{r}
x <- X92st_list |> 
  dplyr::select(St_Name, PR, Lat, Long) |> 
  dplyr::rename(Station = St_Name) |> 
  dplyr::left_join(x, by = "Station")
x
```

```{r}
wmap <- map_data("world")
ca <- wmap |> dplyr::filter(region == "Canada")
mp92 <- ca |> ggplot(aes(long, lat)) +
  geom_path(aes(x = long, y = lat, group = group),
                color = "#737373") +
  geom_polygon(aes(group = group, subgroup = subregion),
                   fill = "#f0f0f0", color = "#525252")+
  scale_color_brewer(palette = "Dark2") +
      theme_void() +
  geom_point(
        data = x, 
        aes(x = Long, y = Lat,# shape = N_modes,
            fill = N_modes, label = Station),
        size = 3, shape = 21
      )+
  labs(fill = "Number of pics")
mp92
ggsave("Figures/fig309.pdf", width = 7, height = 5)
```

```{r}
ggplotly(mp92)
```

```{r}
selected_stations <- x[x$N_modes == 2,]$Station

modes_locations_df <- matrix(
  nrow = length(selected_stations),
  ncol = 3
) |> as.data.frame()
colnames(modes_locations_df) <- c("Station", "Mode 1", "Mode 2")
modes_locations_df$Station <- selected_stations
for (st in selected_stations) {
  print(paste("---------------", st, "----------------"))
  
  df_rescaled_list <- 
      Data_list[[st]] |> 
      rescale()
  df_max <- df_rescaled_list$Max
  df_min <- df_rescaled_list$Min
  
  modes <- df_rescaled_list$Data$Temp |> 
    multimode::locmodes(mod0 = 2)
  modes_loc <- modes$locations[-2]
  modes_loc <-
    (modes_loc * (df_max - df_min)) + df_min
  
  modes_locations_df[
    modes_locations_df$Station == st, -1
  ] <- modes_loc
}
modes_locations_df <- 
  X92st_list |> 
  dplyr::select(St_Name, PR, Lat, Long) |> 
  dplyr::rename(Station = St_Name) |> 
  dplyr::left_join(modes_locations_df, by = "Station")
```

```{r}
ca |> ggplot(aes(long, lat)) +
  geom_path(aes(x = long, y = lat, group = group),
                color = "#737373") +
  geom_polygon(aes(group = group, subgroup = subregion),
                   fill = "#f0f0f0", color = "#525252")+
  scale_color_brewer(palette = "Dark2") +
      theme_void() +
  geom_point(
        data = modes_locations_df, 
        aes(x = Long, y = Lat,# shape = N_modes,
            fill = `Mode 2`, label = Station),
        size = 3, shape = 21
      )
```

```{r}
df_modes <- matrix(nrow = 0, ncol = 6) |> as.data.frame()
colnames(df_modes) <- c("Station", "Year", "Mode_1", "Mode_2", "fvalue_1", "fvalue_2")
for (st in selected_stations) {
  print(paste("---------------", st, "----------------"))
  df_data <- Data_list[[st]]
  yrs <- df_data$Year |> unique()
  df_yr_modes <- data.frame(
    Station = rep(st, length(yrs)),
    Year = yrs,
    Mode_1 = rep(NA, length(yrs)),
    Mode_2 = rep(NA, length(yrs)),
    fvalue_1 = rep(NA, length(yrs)),
    fvalue_2 = rep(NA, length(yrs))
  )
  for (yr in yrs) {
    print(yr)
    df_data_yr <- df_data$Temp[df_data$Year == yr]
    modes <- df_data_yr |> multimode::locmodes(mod0 = 2)
    modes_locs <- modes$locations[-2]
    modes_fval <- modes$fvalue[-2]
    df_yr_modes[df_yr_modes$Station == st &
                  df_yr_modes$Year == yr,c(3:4)] <- modes_locs
    df_yr_modes[df_yr_modes$Station == st &
                  df_yr_modes$Year == yr,c(5:6)] <- modes_fval
  }
  df_modes <- df_modes |> rbind(df_yr_modes)
}
```

```{r}
df_modes_statisctics <-
  df_modes |> 
  group_by(Station) |> 
  summarise(
    mean_M_1 = mean(Mode_1),
    mean_M_2 = mean(Mode_2),
    gap_M_1_M_2 = mean(Mode_2) - mean(Mode_1),
    max_M_1 = mean(Mode_1) + (sqrt(var(Mode_1))/2),
    min_M_1 = mean(Mode_1) - (sqrt(var(Mode_1))/2),
    max_M_2 = mean(Mode_2) + (sqrt(var(Mode_2))/2),
    min_M_2 = mean(Mode_2) - (sqrt(var(Mode_2))/2),
    mean_F_1 = mean(fvalue_1),
    mean_F_2 = mean(fvalue_2),
    max_F_1 = mean(fvalue_1) + (sqrt(var(fvalue_1))/2),
    min_F_1 = mean(fvalue_1) - (sqrt(var(fvalue_1))/2),
    max_F_2 = mean(fvalue_2) + (sqrt(var(fvalue_2))/2),
    min_F_2 = mean(fvalue_2) - (sqrt(var(fvalue_2))/2)
   )
df_modes_statisctics <-
  df_modes_statisctics |> 
  dplyr::mutate(
    M_1_tresh = ifelse(
      mean_M_1 <= 2 & mean_M_1 >= -2,
      "Yes",
      "No"
    ),
    M_2_tresh = ifelse(
      mean_M_2 <= 2 & mean_M_2 >= -2,
      "Yes",
      "No"
    )
  )

X92 <- X92st_list
X92$Region <- ""
X92[X92$PR == "BC", "Region"] <- 
  "British Columbia"
X92[X92$PR == "AB" | X92$PR == "SK" | X92$PR == "MB", "Region"] <- 
  "Prairies"
X92[X92$PR == "ON", "Region"] <- 
  "Ontario"
X92[X92$PR == "QC", "Region"] <- 
  "Quebec"
X92[X92$PR == "NB" | X92$PR == "PE" | X92$PR == "NS" | X92$PR == "NL", "Region"] <- 
  "Atlantic"
X92[X92$PR == "YK" | X92$PR == "NT" | X92$PR == "NU", "Region"] <- 
  "Northern"
X92st_list <- X92
df_modes_statisctics <- 
  X92st_list |> 
  dplyr::mutate(Station = St_Name) |> 
  dplyr::select(Station, PR, Region, Lat, Long) |> 
  merge(
    df_modes_statisctics,
    by = "Station"
  )

df_modes_statisctics |> head()
```

```{r}
wmap <- map_data("world")
ca <- wmap |> dplyr::filter(region == "Canada")
mp92 <- ca |> ggplot(aes(long, lat)) +
  geom_path(aes(x = long, y = lat, group = group),
            color = "black") +
  geom_polygon(aes(group = group, subgroup = subregion),
               fill = "#e0e0e0", color = "black")

fig.a <-
  mp92 +
  geom_point(
    data = df_modes_statisctics, 
    aes(x = Long, y = Lat, color = mean_M_1, shape = M_1_tresh, label = Station),
    size = 3.5
  )  +
  labs(
    #title = "Mode 1",
    shape = "T between -2 and 2(B0C) ?",
    colour = "T(B0C)"
    # size = "??cart-type"
  )  +
  #paletteer::scale_fill_paletteer_c("viridis::plasma") +
  scale_color_continuous(type = "viridis",
                         limits = c(-40, 20))+
  theme_void()

fig.b <-
  mp92 +
  geom_point(
    data = df_modes_statisctics, 
    aes(x = Long, y = Lat, color = mean_M_2, shape = M_2_tresh, label = Station),
    size = 3.5
  )  +
  labs(
    #title = "Mode 1",
    shape = "T between -2 and 2(B0C) ?",
    colour = "T(B0C)"
    # size = "??cart-type"
  )  +
  #paletteer::scale_fill_paletteer_c("viridis::plasma") +
  scale_color_continuous(type = "viridis",
                         limits = c(-30, 20))+
  theme_void()

fig.c <-
  mp92 +
  geom_point(
    data = df_modes_statisctics, 
    aes(x = Long, y = Lat, color = gap_M_1_M_2, label = Station),
    size = 3.5
  )  +
  labs(
    #title = "Mode 1",
    colour = "T(B0C)"
    # size = "??cart-type"
  )  +
  #paletteer::scale_fill_paletteer_c("viridis::plasma") +
  scale_color_continuous(type = "viridis",
                         limits = c(0, 40))+
  theme_void()

fig.a
ggplotly(fig.a)
fig.b
ggplotly(fig.b)
fig.c
ggplotly(fig.c)
```

```{r}
fig_m1_m2 <- 
  df_modes_statisctics |> 
  ggplot(aes(x = mean_M_1, y = mean_M_2, labels = Station, color = Region)) +
  geom_point(size = 2) +
  geom_vline(xintercept = c(-2, 2), linetype=2) +
  geom_hline(yintercept = c(-2, 2), linetype=2) +
  geom_segment(aes(
    x = min_M_1, y = mean_M_2, xend = max_M_1, yend = mean_M_2
  ), linewidth = .5) +
  geom_segment(aes(
    x = mean_M_1, y = min_M_2, xend = mean_M_1, yend = max_M_2
  ), linewidth = .5) +
  # stat_cor(method="pearson")
  labs(
    #title = "Mode 1 vs Mode 2",
    x = "Mode 1", #TeX(r'($M_1$)'),
    y = "Mode 2" #TeX(r'($M_2$)')
  ) +
  theme_classic() +
  scale_color_brewer(palette="Dark2")

fig_m1_m2
ggsave("Figures/fig310_2.pdf", width = 7, height = 5)
```

```{r}
library(ggpubr)
g1 <- ggarrange(fig.a, fig.b,
                labels = c("(a) Mode 1", "(b) Mode 2"), hjust = 0)
g1

ggarrange(g1, fig.c, ncol = 1,
          heights = c(6, 13),
          widths = c(9, 9),
          labels = c("", "(c) Mode2 - Mode1"))
ggsave("Figures/fig310_1.pdf", width = 7, height = 7)
```

```{r}
ggarrange(g1, fig.c, fig_m1_m2, ncol = 1,
          heights = c(6, 13, 15),
          widths = c(9, 9, 9),
          labels = c("", "(c) Mode2 - Mode1", "(d) Mode1 vs Mode2"))
ggsave("Figures/fig311.pdf", width = 7, height = 10)
```

```{r}
# 
```

